Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-07-20 submission/Sze_Obesity_mBio_2016.tex
Incorporated @marcsze's comments (Issue #20) and final editing

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support for the
widely accepted hypothesis that changes in the microbiome are associated
with obesity. This hypothesis was reconsidered by increasing the number
of datasets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status. They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that only one of the studies had sufficient power to detect
a 5\% difference in diversity. When Random Forest machine learning
models were trained on one dataset and then tested using the other 9
datasets, the median accuracy varied between 33.01 and 64.77\%
(median=56.68\%). Although there was support for a relationship between
the microbial communities found in human feces and obesity status, this
association is relatively weak and its detection is confounded by large
interpersonal variation and insufficient sample sizes.

@@

\textbf{\emph{Literature Review and Study Inclusion.}} To perform a
robust meta-analysis and limit inclusion bias, we followed the Preferred
Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA)
guidelines to identify the studies that we analyzed (14). A detailed
description of our selection process and the exact search terms are
provided in the Supplemental Text and in Figure 1. Briefly, we searched
PubMed for original research studies that involved studying obesity and
the human microbiome. The initial search yielded 187 studies. We
identified ten additional studies that were not designed to explicitly
test for an association between the microbiome and obesity. We then
manually curated the 196 studies to select those studies that included
BMI and 16S rRNA gene sequence data. This yielded 10 eligible studies.
An additional study was removed from our analysis because no individuals
in the study had a BMI over 30. Among the final 10 studies, 3 were from
identified from our PubMed search (10, 15, 16), 5 were originally
identified from the 10 studies that did not explicitly investigate
obesity but included BMI data (17--21), and two datasets were used (22,
23) because these publications did not specifically look for any
metabolic or obesity conditions but had control populations and enabled
us to help mitigate against publication biases associated with the
bacterial microbiome and obesity. The ten studies are summarized in
Table 1. For comparison, two of these studies were included in the
Finucane re-analysis study (10, 21) and four of these studies were
included in the Walters re-analysis study (10, 15, 20, 21). The 16S rRNA
gene sequence data from each study was re-analyzed using a similar
approach based on previously described methods for reducing the number
of chimeric sequences and sequencing errors for 454 and Illumina MiSeq
data (24, 25). The sequences were either clustered into operational
taxonomic units (OTUs) using the average neighbor approach (26) or
taxonomic groupings based on their classification using a naive Bayesian
classifier (27).

@@

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for significant differences between the structure of
non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
because the samples are arrayed in a non-dimensional space. nor perform
a pooled analysis using studies that had non-overlapping 16S rRNA gene
sequence regions it is unclear whether these differences reflect a
broader, but perhaps small, shift in community structure between
non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being non-obese or obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the observed AUC values for the OTU-based
models varied between 0.52 and 0.69 indicating a relatively poor ability
to classify individuals (Figure 4A). To test models on other datasets,
we trained models using genus-level phylotype data for each dataset. The
cross-validated AUC values for the models applied to the training
datasets varied between 0.51 and 0.65, again indicating a relatively
poor ability to classify individuals from the original dataset (Figure
4B). For each model we identified the probability where the sum of the
sensitivity and specificity was the highest. We then used this
probability to define a threshold for calculating the accuracy of the
models when applied to the other nine datasets (Figure 5). Although
there was considerable variation in accuracy values for each model, the
median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
(HMP) (median=0.57). We built similar models using taxonomic
representation based on phylum, class, order, and family assignments and
saw no variation in the results (Figure S4). We also attempted to
predict individual BMI values as continuous variables based on the
relative abundance of OTUs and genera. The median percent of the
variance explained with the resulting models was 0.13 for the OTU-based
models and 0.082 for the genus-based models. When we considered the
number of samples, balance of non-obese and obese individuals, and
region within the 16S rRNA gene it was not possible to identify factors
that predictably affected model performance. The ability to predict
obesity status using relative abundance data from the communities is
only marginally better than random. These results suggest that given the
large diversity of microbiome compositions it is difficult to identify a
taxonomic signal that can be associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, and unbalanced
sampling of non-obese and obese individuals. To assess this, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S5-S10). Although there is no biological
rationale for these effect sizes, they represent a range that includes
effect sizes that would be generally considered to be biologically
significant. Only the Goodrich study had power greater than 0.80 to
detect a 5\% difference in Shannon diversity and six of the studies had
enough power to detect a 10\% difference (Figure 6). None of the studies
had sufficient power to detect a 15\% difference between B:F values
(Figure S5). In fact, the maximum power among any of the studies to
detect a 15\% difference in B:F values was 0.25. Among the tests for
relative risk, none of the studies had sufficient power to detect a
Cohen's d of 0.10 and only two studies had sufficient power to detect a
Cohen's d of 0.15. We next estimated how many individuals would need to
have been sampled to have sufficient power to detect the four effect
sizes assuming the observed interpersonal variation from each study and
balanced sampling between the two groups. To detect a 1, 5, 10, or 15\%
difference in Shannon index, the median required sampling effort per
group was approximately 3,400, 140, 35, or 16 individuals, respectively.
To detect a 1, 5, 10, and 15\% difference in B:F values, the median
required sampling effort per group was approximately 160,000, 6,300,
1,600, or 700 individuals, respectively. To detect a 1, 5, 10, and 15\%
difference in relative risk values using Shannon diversity, the median
required sampling effort per group was approximately 39,000, 1,500, 380,
or 170 individuals, respectively. These estimates indicate that most
microbiome studies are underpowered to detect modest effect sizes using
either metric. In the case of obesity, the studies were underpowered to
detect the 0.90 to 6\% difference in diversity that was observed across
the studies.

@@

Alpha diversity metrics are attractive because they distill a complex
dataset to a single value. For example, diversity is a measure of the
entropy in a community and integrates richness and evenness information.
Two communities with little taxonomic similarity can have the same
diversity. Among ecologists the relevance of these metrics is questioned
because it is difficult to ascribe a mechanistic interpretation to their
relationship with stability or disease. Regardless, the concept of a
biologically significant effect size needs to be developed among
microbiome researchers. Alternative metrics could include the ability to
detect a defined difference in the relative abundance of an OTU
representing a defined relative abundance. What makes for a biologically
significant difference or relative abundance is an important point that
has yet to be discussed in the microbiome field. The use of
operationally defined effect sizes should be adequate until it is
possible to decide upon an accepted practice.

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. Some factors such as relationship
between subjects could potentially decrease the amount variation (6) and
other factors such as whether one lives in a rural environment could
increase the amount of variation (28). It was interesting that although
there were ranges in variation across the studies included in our
analysis, the sample size calculations yielded similar results. In
addition, the common experimental designs limit their power. As we
observed, most of the studies included in our analysis were unbalanced
for the variable that we were interested in. This was also true of those
studies that originally sought to identify associations with obesity.
Even with a balanced design, we showed that it was necessary to obtain
approximately 140 and 6,300 samples per group to detect a 5\% difference
in Shannon diversity or B:F, respectively. It was interesting that these
sample sizes agreed across studies regardless of their sequencing
method, region within the 16S rRNA gene, or subject population (Figure
6). This suggests that regardless of the treatment or category, these
sample sizes represent a good starting point for subject recruitment
when using stool samples. Unfortunately, few studies have been published
with this level of subject recruitment. This is troubling since the
positive predictive rate of a significant finding in an underpowered
study is small leading to results that cannot be reproduced (29). Future
microbiome studies should articulate the basis for their experimental
design.

Two previous re-analysis studies have stated that there was not a
consistent association between alpha diversity and obesity (8, 9);
however, neither of these studies made an attempt to pool the existing
data together to try and harness the additional power that this would
give and they did not assess whether the studies were sufficiently
powered to detect a difference. Additionally, our analysis used 16S rRNA
gene sequence data from ten studies whereas the Finucane study used 16S
rRNA gene sequence data from three studies (7, 10, 21) and a metagenomic
study (30) and the Walters study used 16S rRNA gene sequence data from
five studies (10, 15, 20, 21, 28); two studies were included in both
analyses (10, 21). Our analysis included four of these studies (10, 15,
20, 21) and excluded three of the studies because they were too small
(7), only utilized metagenomic data (30), or used short single read
Illumina HiSeq data that has a high error rate making it untractable for
\emph{de novo} OTU clustering (28). The additional seven datasets were
published after the two reviews were performed and include datasets with
more samples than were found in the original studies. Our collection of
ten studies allowed us to largely use the same sequence analysis
pipeline for all datasets and relied heavily on the availability of
public data and access to metadata that included variables beyond the
needs of the original study. To execute this analysis, we created an
automated data analysis pipeline, which can be easily updated to add
additional studies as they become available
(\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}). Similarly,
it would be possible to adapt this pipeline to other body sites and
treatment or variables (e.g.~subject's sex or age).

>>>>

Two recent studies have re-analyzed published data and found that when
datasets are analyzed independently there was limited support for the
widely accepted hypothesis that changes in the microbiome are associated
with obesity. This hypothesis was reconsidered by increasing the number
of datasets and pooling the results across the individual datasets. The
Preferred Reporting Items for Systematic Reviews and Meta-Analyses
(PRISMA) guidelines were applied to identify 10 studies for an updated
and more synthetic analysis. Alpha diversity metrics and the relative
risk of obesity based on those metrics were used to identify a limited
number of significant associations with obesity; however, when the
results of the studies were pooled using a random effects model
significant associations were observed between Shannon diversity, number
of observed OTUs, and Shannon evenness and obesity status. They were not
observed for the ratio of \emph{Bacteroidetes} and \emph{Firmicutes} or
their individual relative abundances. Although these tests yielded small
P-values, the difference between the Shannon diversity index of
non-obese and obese individuals was 2.07\%. A power analysis
demonstrated that only one of the studies had sufficient power to detect
a 5\% difference in diversity. When Random Forest machine learning
models were trained on one dataset and then tested using the other 9
datasets, the median accuracy varied between 33.01 and 64.77\%
(median=56.68\%). Although there was support for a relationship between
the microbial communities found in human feces and obesity status, this
association was relatively weak and its detection is confounded by large
interpersonal variation and insufficient sample sizes.

@@

\textbf{\emph{Literature Review and Study Inclusion.}} To perform a
robust meta-analysis and limit inclusion bias, we followed the Preferred
Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA)
guidelines to identify the studies that we analyzed (14). A detailed
description of our selection process and the exact search terms are
provided in the Supplemental Text and in Figure 1. Briefly, we searched
PubMed for original research studies that involved studying obesity and
the human microbiome. The initial search yielded 187 studies. We
identified 10 additional studies that were not designed to explicitly
test for an association between the microbiome and obesity. We then
manually curated the 197 studies to select those studies that included
BMI and 16S rRNA gene sequence data. This yielded 11 eligible studies.
An additional study was removed from our analysis because no individuals
in the study had a BMI over 30. Among the final 10 studies, 3 were
identified from our PubMed search (10, 15, 16), 5 were originally
identified from the 10 studies that did not explicitly investigate
obesity but included BMI data (17--21), and two datasets were used (22,
23) because these publications did not specifically look for any
metabolic or obesity conditions but had control populations and enabled
us to help mitigate against publication biases associated with the
bacterial microbiome and obesity. The ten studies are summarized in
Table 1. For comparison, two of these studies were included in the
Finucane re-analysis study (10, 21) and four of these studies were
included in the Walters re-analysis study (10, 15, 20, 21). The 16S rRNA
gene sequence data from each study was re-analyzed using a similar
approach based on previously described methods for reducing the number
of chimeric sequences and sequencing errors for 454 and Illumina MiSeq
data (24, 25). The sequences were clustered into operational taxonomic
units (OTUs) using the average neighbor approach (26) and into taxonomic
groupings based on their classification using a naive Bayesian
classifier (27).

@@

\textbf{\emph{Beta diversity analysis.}} Following the approach used by
the Walters and Finucane re-analysis studies, for each dataset we
calculated a Bray-Curtis distance matrix to measure the difference in
the membership and structure of the individuals from each study. We then
used AMOVA to test for significant differences between the structure of
non-obese and obese individuals (Table 1). The Escobar, Goodrich, and
Turnbaugh datasets indicated a significant difference in community
structure (all P\textless{}0.05). Because it was not possible to
ascertain the directionality of the difference in community structure
because the samples are arrayed in a non-dimensional space or perform a
pooled analysis using studies that had non-overlapping 16S rRNA gene
sequence regions, it is unclear whether these differences reflect a
broader, but perhaps small, shift in community structure between
non-obese and obese individuals.

\textbf{\emph{Development of a microbiome-based classifier of obesity.}}
The Walters re-analysis study suggested that it was possible to classify
individuals as being non-obese or obese based on the composition of
their microbiota. We repeated this analysis with additional datasets
using OTU and genus-level phylotype data. For each study we developed a
Random Forest machine learning model to classify individuals. Using
ten-fold cross validation, the cross-validated AUC values for the
OTU-based models varied between 0.52 and 0.69 indicating a relatively
poor ability to classify individuals (Figure 4A). To test models on
other datasets, we trained models using genus-level phylotype data for
each dataset. The cross-validated AUC values for the models applied to
the training datasets varied between 0.51 and 0.65, again indicating a
relatively poor ability to classify individuals from the original
dataset (Figure 4B). For each model we identified the probability where
the sum of the sensitivity and specificity was the highest. We then used
this probability to define a threshold for calculating the accuracy of
the models when applied to the other nine datasets (Figure 5). Although
there was considerable variation in accuracy values for each model, the
median accuracy for each model varied between 0.33 (Turnbaugh) and 0.65
(HMP) (median=0.57). We built similar models using taxonomic
representation based on phylum, class, order, and family assignments and
saw no improvement in the results (Figure S4). We also attempted to
predict individual BMI values as continuous variables based on the
relative abundance of OTUs and genera. The median percent of the
variance explained with the resulting models was 12.9\% for the
OTU-based models and 8.2\% for the genus-based models. When we
considered the number of samples, balance of non-obese and obese
individuals, and region within the 16S rRNA gene for each study it was
not possible to identify factors that predictably affected model
performance. The ability to predict obesity status using relative
abundance data from the communities was only marginally better than
random. These results suggest that given the large diversity of
microbiome compositions it is difficult to identify a taxonomic signal
that can be associated with obesity.

\textbf{\emph{Power and Sample Size Estimate Simulations.}} The
inability to detect a difference between non-obese and obese individuals
could be due to the lack of a true effect or because the study had
insufficient statistical power to detect a difference because of
insufficient sampling, large interpersonal variation, or unbalanced
sampling of non-obese and obese individuals. To assess these factors, we
calculated the power to detect differences of 1, 5, 10, and 15\% in each
of the alpha diversity metrics using the sample sizes used in each of
the studies (Figures 6, S5-S10). Although there is no biological
rationale for these effect sizes, they represent a range that includes
effect sizes that would be generally considered to be biologically
significant. Only the Goodrich study had power greater than 0.80 to
detect a 5\% difference in Shannon diversity and six of the studies had
enough power to detect a 10\% difference (Figure 6A). None of the
studies had sufficient power to detect a 15\% difference between B:F
values (Figure S5). In fact, the maximum power among any of the studies
to detect a 15\% difference in B:F values was 0.25. Among the tests for
relative risk, none of the studies had sufficient power to detect a
Cohen's d of 0.10 and only two studies had sufficient power to detect a
Cohen's d of 0.15. We next estimated how many individuals would need to
have been sampled to have sufficient power to detect the four effect
sizes assuming the observed interpersonal variation from each study and
balanced sampling between the two groups (Figure 6B). To detect a 1, 5,
10, or 15\% difference in Shannon diversity, the median required
sampling effort per group was approximately 3,400, 140, 35, or 16
individuals, respectively. To detect a 1, 5, 10, and 15\% difference in
B:F values, the median required sampling effort per group was
approximately 160,000, 6,300, 1,600, or 700 individuals, respectively.
To detect a 1, 5, 10, and 15\% difference in relative risk values using
Shannon diversity, the median required sampling effort per group was
approximately 39,000, 1,500, 380, or 170 individuals, respectively.
These estimates indicate that most microbiome studies are underpowered
to detect modest effect sizes using either metric. In the case of
obesity, the studies were underpowered to detect the 0.90 to 6\%
difference in diversity that was observed across the studies.

@@

Alpha diversity metrics are attractive because they distill a complex
dataset to a single value. For example, Shannon diversity is a measure
of the entropy in a community and integrates richness and evenness
information. Two communities with little taxonomic similarity can have
the same diversity. Among ecologists the relevance of these metrics is
questioned because it is difficult to ascribe a mechanistic
interpretation to their relationship with stability or disease.
Regardless, the concept of a biologically significant effect size needs
to be developed among microbiome researchers. Alternative metrics could
include the ability to detect a defined difference in the relative
abundance of an OTU representing a defined relative abundance. What
makes for a biologically significant difference or relative abundance is
an important point that has yet to be discussed in the microbiome field.
The use of operationally defined effect sizes should be adequate until
it is possible to decide upon an accepted practice.

By selecting a range of possible effect sizes, we were able to
demonstrate that most studies are underpowered to detect modest
differences in alpha diversity metrics and phylum-level relative
abundances. Several factors interact to limit the power of microbiome
studies. There is wide interpersonal variation in the diversity and
structure of the human microbiome. Some factors such as relationship
between subjects could potentially decrease the amount variation (6) and
other factors such as whether one lives in a rural environment could
increase the amount of variation (28). In addition, the common
experimental designs limit their power. As we observed, most of the
studies included in our analysis were unbalanced for the variable that
we were interested in. This was also true of those studies that
originally sought to identify associations with obesity. Even with a
balanced design, we showed that it was necessary to obtain approximately
140 and 6,300 samples per group to detect a 5\% difference in Shannon
diversity or B:F, respectively. It was interesting that these sample
sizes agreed across studies regardless of their sequencing method,
region within the 16S rRNA gene, or subject population (Figure 6). This
suggests that regardless of the treatment or category, these sample
sizes represent a good starting point for subject recruitment when using
stool samples. Unfortunately, few studies have been published with this
level of subject recruitment. This is troubling since the positive
predictive rate of a significant finding in an underpowered study is
small leading to results that cannot be reproduced (29). Future
microbiome studies should articulate the basis for their experimental
design.

Two previous re-analysis studies have stated that there was not a
consistent association between alpha diversity and obesity (8, 9);
however, neither of these studies made an attempt to pool the existing
data together to try and harness the additional power that this would
give and they did not assess whether the studies were sufficiently
powered to detect a difference. Additionally, our analysis used 16S rRNA
gene sequence data from ten studies whereas the Finucane study used 16S
rRNA gene sequence data from three studies (7, 10, 21) and a metagenomic
study (30) and the Walters study used 16S rRNA gene sequence data from
five studies (10, 15, 20, 21, 28); two studies were included in both
analyses (10, 21). Our analysis included four of these studies (10, 15,
20, 21) and excluded three of the studies because they were too small
(7), only utilized metagenomic data (30), or used short single read
Illumina HiSeq data that has a high error rate making it intractable for
\emph{de novo} OTU clustering (28). The additional seven datasets were
published after the two reviews were performed and include datasets with
more samples than were found in the original studies. Our collection of
ten studies allowed us to largely use the same sequence analysis
pipeline for all datasets and relied heavily on the availability of
public data and access to metadata that included variables beyond the
needs of the original study. To execute this analysis, we created an
automated data analysis pipeline, which can be easily updated to add
additional studies as they become available
(\url{https://github.com/SchlossLab/Sze_Obesity_mBio_2016/}). Similarly,
it would be possible to adapt this pipeline to other body sites and
treatment or variables (e.g.~subject's sex or age).
